---
title: "1974...bird attacks?"
author: "Allison Horst"
format: html
execute: 
  warning: false
  message: false
editor: visual
---

## A-well a bird bird bird...

Why did this post happen? Well, at around 11pm last night I saw [this tweet](https://twitter.com/tcarpenter216/status/1580771753873518593):

\
[![Twitter post with an image of two tables, together titled "Bird Attacks 1974." The two tables are "Table 1. Bird attacks: body parts injured" and "Table 2. Bird attacks: nature of wounds." The tables contain counts for various injuries and wounds . The data is silly and fake. For example, some of the "wounds" in the dataset include "Cut by stinging remarks" from a bird, and "Feelings hurt" by a bird.](images/paste-68076610.png){width="553"}](https://twitter.com/tcarpenter216/status/1580771753873518593)

Tom's instinct makes perfect sense. My first thought was ["CHICKEN WINGS."](https://www.insider.com/restaurants-pricing-confusing-math-2018-10) So I spent a couple sleepless hours with this weird (and obviously fake) data set about bird attacks in 1974, because it felt like a good October distraction and an opportunity to get some data wrangling rust off.

In this post:

-   Image cropping with magick

```{=html}
<!-- -->
```
-   Optical character recognition (OCR) with tesseract

-   Low elegance, no shame data tidying in R

-   Static & interactive visualizations in JavaScript

fowl, cuckoo, wing, fly, flap, take-off, beak, feather, soar, flock, rail, lark, duck, swallow

## Get the tools and data:

Attach libraries:

```{r}
library(magick)
library(tesseract)
library(tidyverse)
library(patchwork)
library(kableExtra)
```

Read in and crop the image with magick:

```{r}
birds_img <- image_read("bird_attacks_tweet.jpg")
birds <- image_crop(birds_img, "1100x1100+50+200")
```

Here's the cropped version of the image:

```{r, fig.alt="Cropped version of the bird attacks image, now containing only the tables and title."}
#| echo: false
birds
```

## Optical character recognition 

Optical character recognition (OCR), or text recognition, lets you extract text information from images (e.g. non-selectable PDFs, JPEGs, PNGs, etc.). The [tesseract package](https://cran.r-project.org/web/packages/tesseract/vignettes/intro.html) in R "provides R bindings to [Tesseract](https://github.com/tesseract-ocr/tesseract): a
powerful optical character recognition (OCR) engine that supports over 100 languages."

Create the engine, then OCR those birds:

```{r}
ocr_eng <- tesseract("eng")
text <- tesseract::ocr(birds, engine = ocr_eng)
```

Take a look at the extracted text. I will never stop thinking this is amazing.

```{r}
cat(text)
```

## Cleaning

Here's my "no regerts" code for data cleaning. I would love to see other ways that people wrangle these birds.

```{r}
# Separating into columns
bird_dat <- data.frame(text) |> 
  mutate(text = strsplit(as.character(text), "\n")) |>  
    unnest(text) |> 
  filter(text != "") |>
  slice(-c(1:2)) |> 
  separate(col = "text" , 
           into = c("body_parts", "leftover"), 
           sep = "(?<=[a-zA-Z])\\s*(?=[0-9])",
           extra = "merge") |> 
  separate(col = "leftover",
           into = c("body_parts_count", "leftover"),
           sep = " ",
           extra = "merge") |> 
  separate(col = "leftover" , 
           into = c("nature_of_wounds", "nature_of_wounds_counts"), 
           sep = "(?<=[a-zA-Z])\\s*(?=[0-9])",
           extra = "merge") |> 
  slice(-1)

# Some artisanal hacking. Not proud, but tired... 
bird_dat$nature_of_wounds[16] <- "Body not recovered"
bird_dat$nature_of_wounds_counts[16] <- bird_dat$body_parts_count[16]
bird_dat$body_parts[16] <- NA
bird_dat$body_parts_count[16] <- NA

# At this point, splitting into separate data frames:
body_parts_injured <- data.frame(body_parts = bird_dat$body_parts, n = bird_dat$body_parts_count) |> 
  drop_na()

wounds <- data.frame(wound_nature = bird_dat$nature_of_wounds, 
                     n = bird_dat$nature_of_wounds_counts)

# Where nature_of_wounds_counts is NA, combine text w/ following row.
# No, this does not follow any cohesive data cleaning philosophy.
for (i in 1:(length(wounds$n) - 1)) {
  if (is.na(wounds$n[i])) {
    wounds$wound_nature[i+1] <- paste(wounds$wound_nature[i], wounds$wound_nature[i+1])
    }
}

# Then get rid of the duplicate text rows still containing NA
wounds <- wounds |> 
  drop_na() |> 
  mutate(n = as.numeric(n))

```

Which gives us these two data frames:

```{r}
#| layout-ncol: 2
#| echo: false

body_parts_injured |> 
  kable() |> 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

wounds |> 
  kable() |> 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)

```

## Viz those birds

I'm going to do some JavaScript data visualization in OJS cells, which I can do [right in this Quarto doc](https://quarto.org/docs/interactive/ojs/ojs-cells.html) 🤯 Curious about adding more JavaScript to your Quarto world? I highly recommend [Sharon Machlis](https://twitter.com/sharon000)' recent series, [*A beginner's guide to using Observable JavaScript, R, and Python with Quarto*](https://www.infoworld.com/article/3674789/a-beginners-guide-to-using-observable-javascript-r-and-python-with-quarto.html).

First, I'll make my R objects available for use in OJS cells:

```{r}
ojs_define(wounds = wounds)
ojs_define(body_parts_injured = body_parts_injured)
```

At this point, I switch over to OJS cells -- code fencing is shown for OJS cells below to clarify the switch from R. First step, transposing the data - which still hurts my column-loving heart a bit, but I'm getting used to it.

```{ojs}
//| echo: fenced
wounds_transpose = transpose(wounds)
injuries_transpose = transpose(body_parts_injured)
```

Then some JavaScript visualizations.

Here, a bar chart created with [Observable Plot](https://observablehq.com/@observablehq/plot). If you're used to working in ggplot2, building visualizations following the grammar of graphics in Plot feels similar.

```{ojs}
//| echo: fenced
Plot.plot({
  marks: [
    Plot.barX(wounds_transpose, 
    {y: "wound_nature", x: "n", 
    sort: {y: "x", reverse: true},
    fill: "navy"
    })
  ], 
  marginLeft: 200, 
  x: {label: "Number"},
  y: {label: "Type of injury from bird attack"}
})
```

We know this data is questionable because clearly the two most common bird wounds are "Cut by stinging remarks" and "Feelings hurt." I suspect this may be due to social desirability bias. Further research is needed.

You know what viz I really need (read: I really don't, but keep reading because this is cool anyway)? A D3 [bubble chart](https://observablehq.com/@d3/bubble-chart) of those wounds, with circle area proportional to counts. Instead of building this from scratch, I'm going to use [Observable imports](https://observablehq.com/@observablehq/import) to have it (or any other named thing in a published Observable notebook) at-my-fingertips. Again, this feels like a bit of magic.

```{ojs}
//| echo: fenced
import {BubbleChart} from "@d3/bubble-chart"
```

Now, the D3 BubbleChart function is available for me to usehere. Right in my Quarto doc. I'm clapping.

```{ojs}
//| echo: fenced
BubbleChart(wounds_transpose, 
            {label: (d) => `${d.wound_nature}\n\n${d.n}`,
            value: d => d.n,
            group: d => d.n,
            title: (d) => `${d.wound_nature}\n\n${d.n}`
            })
```

I refuse to leave this without some visualization of the "body parts injured" data. Let's make a (spin the wheel) [D3 DONUT CHART](https://observablehq.com/@d3/donut-chart)!

Again, I import from th Observable Notebook (notice the slug is \\@d3/donut-chart, and the named function is DonutChart):

```{ojs}
import {DonutChart} from "@d3/donut-chart"
```

With the `DonutChart` function now available for me to use here, I can make my plot:

```{ojs}

DonutChart(injuries_transpose, {
  name: d => d.body_parts,
  value: d => d.n,
  width,
  height: 500
})
```

In the spirit of publicly sharing works-in-progress and imperfect things, I'll stop here. I look forward to seeing other ways that people wrangle & visualize data from the 1974 bird attack epidemic. Stay safe out there, friends.
